---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# LTMRdata

<!-- badges: start -->
<!-- badges: end -->

Integration of the Fall Midwater Trawl, Bay Study, and Suisun Marsh Fish Study data for use in the IEP long-term monitoring survey review.

## Installation

You can install `LTMRdata` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("sbashevkin/LTMRdata")
```

# Usage

```{r}
require(LTMRdata)
```

## To access the individual source datasets

Suisun Marsh Fish Study
```{r}
data(Suisun)
str(Suisun)
```

Fall Midwater Trawl
```{r}
data(FMWT)
str(FMWT)
```

Bay Study
```{r}
data(Baystudy)
str(Baystudy)
```

## To access the integrated dataset

```{r}
Data <- LTMRpilot()
str(Data)
```

# Code of Conduct

Please note that the 'LTMRdata' project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
